Measurement and Probability in Fuzzy Quantum Theories

### 2016-11-27 proposal/prop.tex
with andy

\begin{example}[One-qubit quantum probability space] Consider a
  one-qubit Hilbert space with each event interpreted as a possible
  post-measurement state. For example, the event $\proj{0}$ indicates
  that the post-measurement state will be $\ket{0}$; the probability
  of such an event depends on the current state; the event $\proj{1}$
  indicates that the post-measurement state will be $\ket{1}$; the
  event $\proj{\ps}$ where
  $\ket{\ps}=\frac{1}{\sqrt{2}}(\ket{0}+\ket{1})$ indicates that the
  post-measurement state will be $\ket{\ps}$; the event
  $\mathbb{1}=\proj{0}+\proj{1}$ indicates that the post-measurement
  state will be a linear combination of $\ket{0}$ and $\ket{1}$ and is
  clearly certain; finally the empty event $\mathbb{0}$ states that
  the post-measurement state will be the empty state and is
  impossible. As in the classical case, a probability measure is a
  function that maps events to $[0,1]$. Here is a partial
  specification of a possible probability measure that would be
  induced by a system whose current state is $\ket{0}$:
\[
\mu\left(\mathbb{0}\right)=0,\quad
\mu\left(\mathbb{1}\right)=1,\quad
\mu\left(\proj{0}\right)=1,\quad
\mu\left(\proj{1}\right)=0,\quad
\mu\left(\proj{\ps}\right)=1/2,\quad
\ldots
\]
Note that, similarly to the classical case, the probability of
$\mathbb{1}$ is 1 and the probability of collections of orthogonal
events (e.g., $\proj{0}+\proj{1}$) is the sum of the individual
probabilities.  A collection of non-orthogonal events (e.g.,
$\proj{0}$ and $\proj{\ps}$) is however not even a valid event. In the
classical example, we argued that each probability measure is uniquely
determined by two actual coins. A similar (but much more subtle)
argument is valid also in the quantum case. By postulates of quantum
mechanics and Gleason's theorem, it turns out that for large enough
quantum systems, each probability measure is uniquely determined by an
actual quantum state as discussed next.
\end{example}

@@

\noindent A quantum probability measure can be easily constructed if
one knows the current state of the quantum system by using the Born
rule~\cite{Born1983,Mermin2007,Jaeger2007}.  Specifically, for each
pure normalized quantum state $\ket{\phi}$, the Born rule induces a
probability measure $\mu_{\phi}^{B}$ defined as
$\mu_{\phi}^{B}(P)=\ip{\phi}{P\phi}$. The situation generalizes to
mixed states $\rho = \sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where
$\sum_{j=1}^{N}q_{j}=1$ in which case the generalized Born rule
induces a probability measure $\mu_{\rho}^{B}$ defined
as follows~\cite{peres1995quantum,544199,Jaeger2007}:
\[
\mu_{\rho}^{B}\left(P\right) = 
\Tr\left(\rho P\right) = 
\sideset{}{_{j=1}^{N}}\sum q_{j}\mu_{\phi_{j}}^{B}\left(P\right)
\]
Conversely every probability measure must be of this form.

@@

\begin{definition}[Quantum Interval-valued Probability Measure]\label{def:QuantumInterval-valuedProbability}
  Given a Hilbert space $\Hilb$ with quantum events
  (projections)~$\events$, and a collection of
  intervals~$\mathscr{I}$, a \emph{quantum
    $\mathscr{I}$-interval-valued probability measure} is a
  function~$\bar{\mu}:\events\rightarrow\mathscr{I}$ such that:
\begin{itemize}
\item $\bar{\mu}(\mathbb{0})=\left[0,0\right]$. 
\item $\bar{\mu}(\mathbb{1})=\left[1,1\right]$. 
\item For any projection $P$, 
\begin{equation}
\bar{\mu}\left(\mathbb{1}-P\right)=\left[1,1\right]-\bar{\mu}\left(P\right)\textrm{ .}\label{eq:QuantumInterval-valuedProbability-Complement}
\end{equation}
\item For a set of mutually orthogonal projections $\left\{ P_{i}\right\} _{i=1}^{N}$,
we have:
\begin{equation}
\bar{\mu}\left(\sideset{}{_{i=1}^{N}}\sum P_{i}\right)\subseteq\sideset{}{_{i=1}^{N}}\sum\bar{\mu}\left(P_{i}\right)
\label{eq:QuantumInterval-valuedProbability-Inclusion}
\end{equation}
\end{itemize}
\end{definition}

@@

\begin{definition}
  Given a quantum interval-valued probability
  measure~$\bar{\mu}:\events\rightarrow\mathscr{I}$:
\begin{itemize}
\item A \emph{core} of $\bar{\mu}$ is the set
  $\mathrm{core}\left(\bar{\mu}\right)=\left\{
    \pmeas:\events\rightarrow[0,1] ~\middle|~ \forall
    E\in\events.~\pmeas\left(E\right)\in\bar{\mu}\left(E\right)\right\}$.
\item $\bar{\mu}$ is called convex if 
\begin{equation}
  \bar{\mu}\left(P_{0}\vee P_{1}\right)+\bar{\mu}\left(P_{0}P_{1}\right)\subseteq\bar{\mu}\left(P_{0}\right)+\bar{\mu}\left(P_{1}\right)\label{eq:QuantumInterval-valuedProbability-Convex}
\end{equation}
 for all commuting $P_{0},P_{1}\in\events$. 
\end{itemize}
\end{definition}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{measure4.pdf}
\end{center}
\caption{\label{fig:three-dimensional-4-value}This figure illustrates
  cases 1 and 2 of example~\ref{ex:three-dimensional-4-value} plotted
  in $\mathbb{R}^{3}$. The red and green dotted vectors are $\ket{0}$
  and $\ket{\ps}$ respectively.  All possible vectors of the subspaces
  $\ket{0_{\theta,\gamma}^{\perp}}$ and
  $\ket{\ps_{\theta,\gamma}^{\perp}}$ are drawn in the red and green
  circles, respectively. Within the circles, a dotted
  vector~$\ket{\psi}$ means $\bar{\mu}(\proj{\psi})=\likely$;
  otherwise, $\bar{\mu}(\proj{\psi})=\unlikely$. The gray vector is a
  generic vector~$\ket{0_{\theta,\gamma}^{\perp}}$, and the red and
  green solid vectors are normalized
  $\ket{0}\times\ket{0_{\theta,\gamma}^{\perp}}$ and
  $\ket{\ps}\times\ket{0_{\theta,\gamma}^{\perp}}$, respectively.}
\end{figure}

@@

@@

\paragraph*{Publications.} Several papers were published to report on
the results. On the foundational side, we have the following papers:
\begin{itemize}
\item Computing with Semirings and Weak Rig Groupoids, European
  Symposium on Programming, 2016.
\item Superstructural Reversible Logic, International Workshop on
  Linearity, 2014.
\item Theseus: A High Level Language for Reversible
  Computing. Work-in-Progress report in the Conference on Reversible
  Computation, 2014.
\item Discrete Quantum Theories, J. Phys. A: Math. Theor. 2014.
\item Geometry of Discrete Quantum Computing, J. Phys. A:
  Math. Theor. 2013.
\item Isomorphic Interpreters from Logically Reversible Abstract
  Machines. Workshop on Reversible Computation 2012.
\end{itemize}
On the applications side, we have the following papers: 
\begin{itemize}
\item Copredication in Homotopy Type Theory: A homotopical approach
  for formal semantics of natural languages. Workshop on Natural
  Language and Computer Science, 2016.
\item Contract Monitoring Semantics as Patterns of
  Communication. International Conference on Functional Programming,
  2015.
\item Extensible Effects: An Alternative to Monad
  Transformers. Haskell Symposium, 2013.
\item Encoding Secure Information Flow with Restricted Delegation and
  Revocation in Haskell. Workshop on Functional Programming Concepts
  in Domain-Specific Languages, 2013.
\end{itemize}                                                                          
In addition to the publications, Dr. Sabry and collaborators maintain
a public repository with software related to the project. Dr. Sabry
was also invited to give talks at several venues, most notably:
\begin{itemize}
\item Colloquiumfest, University of Saskatchewan, Canada, 2014 
\item Workshop on Quantum Information and Foundations of Quantum
  Mechanics, University of British Columbia, Vancouver, Canada, 2013
\end{itemize}

>>>>

\begin{example}[One-qubit quantum probability space] Consider a
  one-qubit Hilbert space with each event interpreted as a possible
  post-measurement state. For example, the event $\proj{0}$ indicates
  that the post-measurement state will be $\ket{0}$; the probability
  of such an event depends on the current state; the event $\proj{1}$
  indicates that the post-measurement state will be $\ket{1}$; the
  event $\proj{\ps}$ where
  $\ket{\ps}=\frac{1}{\sqrt{2}}(\ket{0}+\ket{1})$ indicates that the
  post-measurement state will be $\ket{\ps}$; the event
  $\mathbb{1}=\proj{0}+\proj{1}$ indicates that the post-measurement
  state will be a linear combination of $\ket{0}$ and $\ket{1}$ and is
  clearly certain; finally the empty event $\mathbb{0}$ states that
  the post-measurement state will be the empty state and is
  impossible. As in the classical case, a probability measure is a
  function that maps events to $[0,1]$. Here is a partial
  specification of a possible probability measure that would be
  induced by a system whose current state is $\ket{0}$,
  $\mu\left(\mathbb{0}\right)=0$, $\mu\left(\mathbb{1}\right)=1$,
  $\mu\left(\proj{0}\right)=1$, $\mu\left(\proj{1}\right)=0$,
  $\mu\left(\proj{\ps}\right)=1/2$, \ldots. Note that, similarly to
  the classical case, the probability of $\mathbb{1}$ is 1 and the
  probability of collections of orthogonal events (e.g.,
  $\proj{0}+\proj{1}$) is the sum of the individual probabilities.  A
  collection of non-orthogonal events (e.g., $\proj{0}$ and
  $\proj{\ps}$) is however not even a valid event. In the classical
  example, we argued that each probability measure is uniquely
  determined by two actual coins. A similar (but much more subtle)
  argument is valid also in the quantum case. By postulates of quantum
  mechanics and Gleason's theorem, it turns out that for large enough
  quantum systems, each probability measure is uniquely determined by
  an actual quantum state as discussed next.
\end{example}

@@

\noindent A quantum probability measure can be easily constructed if
one knows the current state of the quantum system by using the Born
rule~\cite{Born1983,Mermin2007,Jaeger2007}.  Specifically, for each
pure normalized quantum state $\ket{\phi}$, the Born rule induces a
probability measure $\mu_{\phi}^{B}$ defined as
$\mu_{\phi}^{B}(P)=\ip{\phi}{P\phi}$. The situation generalizes to
mixed states $\rho = \sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where
$\sum_{j=1}^{N}q_{j}=1$ in which case the generalized Born rule
induces a probability measure $\mu_{\rho}^{B}$ defined
as~\cite{peres1995quantum,544199,Jaeger2007}
$\mu_{\rho}^{B}\left(P\right) = \Tr\left(\rho P\right) =
\sideset{}{_{j=1}^{N}}\sum q_{j}\mu_{\phi_{j}}^{B}\left(P\right)$.
Conversely every probability measure must be of this form.

@@

\begin{definition}[Quantum Interval-valued Probability Measure]\label{def:QuantumInterval-valuedProbability}
  Given a Hilbert space $\Hilb$ with quantum events
  (projections)~$\events$, and a collection of
  intervals~$\mathscr{I}$, a \emph{quantum
    $\mathscr{I}$-interval-valued probability measure} is a
  function~$\bar{\mu}:\events\rightarrow\mathscr{I}$ such that:
\begin{itemize}
\item $\bar{\mu}(\mathbb{0})=\left[0,0\right]$. 
\item $\bar{\mu}(\mathbb{1})=\left[1,1\right]$. 
\item For any projection $P$,
  $\bar{\mu}\left(\mathbb{1}-P\right)=\left[1,1\right]-\bar{\mu}\left(P\right)$.
\item For a set of mutually orthogonal projections
  $\left\{ P_{i}\right\} _{i=1}^{N}$, we have
  $\bar{\mu}\left(\sideset{}{_{i=1}^{N}}\sum
    P_{i}\right)\subseteq\sideset{}{_{i=1}^{N}}\sum\bar{\mu}\left(P_{i}\right)$.
\end{itemize}
\end{definition}

@@

\begin{definition}
  Given a quantum interval-valued probability
  measure~$\bar{\mu}:\events\rightarrow\mathscr{I}$:
\begin{itemize}
\item A \emph{core} of $\bar{\mu}$ is the set
  $\mathrm{core}\left(\bar{\mu}\right)=\left\{
    \pmeas:\events\rightarrow[0,1] ~\middle|~ \forall
    E\in\events.~\pmeas\left(E\right)\in\bar{\mu}\left(E\right)\right\}$.
\item $\bar{\mu}$ is called convex if
  $\bar{\mu}\left(P_{0}\vee
    P_{1}\right)+\bar{\mu}\left(P_{0}P_{1}\right)\subseteq\bar{\mu}\left(P_{0}\right)+\bar{\mu}\left(P_{1}\right)$
  for all commuting $P_{0},P_{1}\in\events$.
\end{itemize}
\end{definition}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.48]{measure4.pdf}
\end{center}
\caption{\label{fig:three-dimensional-4-value}This figure illustrates
  cases 1 and 2 of example~\ref{ex:three-dimensional-4-value} plotted
  in $\mathbb{R}^{3}$. The red and green dotted vectors are $\ket{0}$
  and $\ket{\ps}$ respectively.  All possible vectors of the subspaces
  $\ket{0_{\theta,\gamma}^{\perp}}$ and
  $\ket{\ps_{\theta,\gamma}^{\perp}}$ are drawn in the red and green
  circles, respectively. Within the circles, a dotted
  vector~$\ket{\psi}$ means $\bar{\mu}(\proj{\psi})=\likely$;
  otherwise, $\bar{\mu}(\proj{\psi})=\unlikely$. The gray vector is a
  generic vector~$\ket{0_{\theta,\gamma}^{\perp}}$, and the red and
  green solid vectors are normalized
  $\ket{0}\times\ket{0_{\theta,\gamma}^{\perp}}$ and
  $\ket{\ps}\times\ket{0_{\theta,\gamma}^{\perp}}$, respectively.}
\end{figure}

@@

\begin{figure}
\begin{center}
  \includegraphics[scale=0.48]{HoughLineData.pdf}
\end{center}
  \caption{Curves of possible line parameters determined by noisy
    data samples on a straight line in 2D space.}
\label{houghline.fig}
\end{figure}

%%%%%
\subsection{Proposed Activity V: Applications}

A classic evidence-accumulation method known as the Hough
transform~\cite{BallardGenHough1981}, which is very closely related to
Bayesian evidence accumulation
methods~\cite{CrandallEtal2005,CrandallEtal2006}, provides another
example of how our proposal to define physical theories without exact
real numbers could be combined with the measurement process. Simply
described, the Hough transform as an abstraction of Bayesian schemes
starts with a model of the theory being measured, typically a model
that would ordinarily be treated {\it after\/} gathering the data by
applying least squares to choose the optimal parameters.  But in this
method, instead of thinking of, say, the model of a straight line
$\hat{\mathbf{n}}\cdot \mathbf{x} = d$ as an equation for the points
on a line $ \mathbf{x}_{i} = (x_{i},y_{i})$, we think of it as a
time-accumulating {\it set\/} of equations for the distance $d$ of the
nearest point on the line and the direction
$ \hat{\mathbf{n}} = (\cos \theta, \sin \theta)$ from the origin to
that point (the direction perpendicular to the line).  Thus each point
of {\it evidence\/}, $(x_{i},y_{i})$, produces an {\it equation in the
  parameter space\/} that is effectively a plot of all the possible
values of $d(\theta)$ as a function of the possible angular directions
$\theta$ pointing to the family of all possible lines.  A typical
result from a set of errorful (it is impossible to measure real
numbers, we recall) data measurements of points on a hypothetical
straight line is shown in Fig.~\ref{houghline.fig}.

The interesting step for us is now to realize that the {\it parameter
  curves\/} shown in the figure do not need, or indeed, must not, be
themselves infinitely precise real numbers, but they must themselves
incorporate in some way a theory consistent with forbidding absolute
precision.  In practice, then, the sinusoidal lines in Figure
\ref{houghline.fig} all become fuzzy, and the accumulation of these
fuzzy distributions produces peaks of likelihood around the values
$(d_0, \theta_0)$ that are the optimal choices for the line describing
the set of data.  Note that even {\it one\/} data point provides
evidence on a set of hypotheses. The simple analysis we have outlined
here is easily extended to much more complex situations in higher
dimensions (see, e.g.,~\cite{BallardGenHough1981}).  Our point here is
that essentially any physical theory that would ordinarily be
described by a mathematical model using infinite precision numbers is
potentially nonsensical from a computational point of view: computer
scientists know perfectly well, and have been dealing with since the
time of Turing and Shannon, that every single bit of precision in any
number has a cost, and that there is not enough energy in the universe
to computer even one general real number.  Thus the combination of
physical models that, in some intrinsic fashion, contain only
computable numbers, with theories of measurement and evidence in the
fashion of the Hough transform example we have just presented, could
contain the outline of entirely new ways of understanding what
mathematical models of the physical world and its measurement actually
could be.

@@

\paragraph*{Publications.} Several papers were published to report on
the results. On the foundational side, we have the following papers:
(1) Computing with Semirings and Weak Rig Groupoids, European
Symposium on Programming, 2016, (2) Superstructural Reversible Logic,
International Workshop on Linearity, 2014, (3) Theseus: A High Level
Language for Reversible Computing. Work-in-Progress report in the
Conference on Reversible Computation, 2014, (4) Discrete Quantum
Theories, J. Phys. A: Math. Theor. 2014., (5) Geometry of Discrete
Quantum Computing, J. Phys. A: Math. Theor. 2013., (6) Isomorphic
Interpreters from Logically Reversible Abstract Machines. Workshop on
Reversible Computation 2012. On the applications side, we have the
following papers: (1) Copredication in Homotopy Type Theory: A
homotopical approach for formal semantics of natural
languages. Workshop on Natural Language and Computer Science, 2016.,
(2) Contract Monitoring Semantics as Patterns of
Communication. International Conference on Functional Programming,
2015., (3) Extensible Effects: An Alternative to Monad
Transformers. Haskell Symposium, 2013., (4) Encoding Secure
Information Flow with Restricted Delegation and Revocation in
Haskell. Workshop on Functional Programming Concepts in
Domain-Specific Languages, 2013. In addition to the publications,
Dr. Sabry and collaborators maintain a public repository with software
related to the project. Dr. Sabry was also invited to give talks at
several venues, most notably the Colloquiumfest, University of
Saskatchewan, Canada, 2014, and the Workshop on Quantum Information
and Foundations of Quantum Mechanics, University of British Columbia,
Vancouver, Canada, 2013
